---
title: "Validating rkainu against published reference values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating rkainu against published reference values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rkainu)
```

## Spirometry (Kainu et al. 2015)

Kainu A, Timonen KL, Toikka J, et al. Reference values of spirometry for
Finnish adults. *Clin Physiol Funct Imaging*. 2016;36(5):346-358.

The `rkainu` package implements the reference equations from this paper,
covering 10 spirometry parameters for Finnish adults aged 18--90.

### Males, 177 cm

```{r spiro_males}
ages <- c(20, 30, 40, 50, 60, 70, 80)
params <- c("FVC", "FEV1", "FEV1FVC", "PEF")

male_pred <- kainu_pred(ages, 177, 1, params)
male_pred <- cbind(age = ages, round(male_pred, 3))
knitr::kable(male_pred, caption = "Predicted values -- males, 177 cm")
```

### Females, 164 cm

```{r spiro_females}
female_pred <- kainu_pred(ages, 164, 2, params)
female_pred <- cbind(age = ages, round(female_pred, 3))
knitr::kable(female_pred, caption = "Predicted values -- females, 164 cm")
```

### FEV1 decline with age

```{r fev1_plot, fig.width=6, fig.height=4}
age_seq <- seq(18, 84, by = 1)
fev1_m <- kainu_pred(age_seq, 177, 1, "FEV1")
fev1_f <- kainu_pred(age_seq, 164, 2, "FEV1")

plot(age_seq, fev1_m, type = "l", lwd = 2, col = "#235787",
     xlab = "Age (years)", ylab = "Predicted FEV1 (L)",
     main = "Kainu et al. (2015) -- Predicted FEV1 vs age",
     ylim = c(1.5, 5.5))
lines(age_seq, fev1_f, lwd = 2, col = "#e63946")
legend("topright", legend = c("Male 177 cm", "Female 164 cm"),
       col = c("#235787", "#e63946"), lwd = 2, bty = "n")
```

### Z-score verification

When the observed value equals the predicted value, the z-score should be
exactly zero:

```{r zscore_check}
pred_fev1 <- kainu_pred(50, 177, 1, "FEV1")
z <- kainu_zscore(50, 177, 1, FEV1 = pred_fev1)
cat("Predicted FEV1:", round(pred_fev1, 4), "L\n")
cat("Z-score when observed = predicted:", z, "\n")
```

### Lower limit of normal

```{r lln}
lln_table <- data.frame(
  age = ages,
  pred_FEV1 = round(kainu_pred(ages, 177, 1, "FEV1"), 3),
  LLN_FEV1 = round(kainu_lln(ages, 177, 1, "FEV1"), 3),
  pred_FVC = round(kainu_pred(ages, 177, 1, "FVC"), 3),
  LLN_FVC = round(kainu_lln(ages, 177, 1, "FVC"), 3)
)
knitr::kable(lln_table, caption = "Predicted vs LLN -- males, 177 cm")
```

### Data.frame workflow

```{r df_example}
patients <- data.frame(
  id = 1:3,
  age = c(35, 55, 72),
  height = c(182, 170, 168),
  sex = c(1, 2, 1),
  FEV1 = c(4.2, 2.5, 2.1),
  FVC = c(5.3, 3.2, 3.0)
)

result <- kainu_df(patients, params = c("FEV1", "FVC"))
knitr::kable(round(result[, -1], 2), caption = "kainu_df() output")
```

## DLCO (Kainu et al. 2017)

Kainu A, Timonen KL, Toikka J, et al. Reference values of diffusing
capacity for Finnish adults. *Clin Physiol Funct Imaging*. 2018;38(3):413-420.

DLCO functions require age, height, weight, and sex. Three parameters are
available: DLCOc (corrected for haemoglobin), DLCOc/VA (transfer coefficient),
and VA (alveolar volume).

### Predicted values

```{r dlco_pred}
dlco_table <- data.frame(
  age = ages,
  DLCO_male = round(dlco_pred(ages, 177, 80, 1, "DLCO"), 2),
  DLCO_female = round(dlco_pred(ages, 164, 65, 2, "DLCO"), 2),
  DLCOVA_male = round(dlco_pred(ages, 177, 80, 1, "DLCOVA"), 2),
  DLCOVA_female = round(dlco_pred(ages, 164, 65, 2, "DLCOVA"), 2),
  VA_male = round(dlco_pred(ages, 177, 80, 1, "VA"), 2),
  VA_female = round(dlco_pred(ages, 164, 65, 2, "VA"), 2)
)
knitr::kable(dlco_table, caption = "Predicted DLCO values by age and sex")
```

### DLCO decline with age

```{r dlco_plot, fig.width=6, fig.height=4}
dlco_m <- dlco_pred(age_seq, 177, 80, 1, "DLCO")
dlco_f <- dlco_pred(age_seq, 164, 65, 2, "DLCO")

plot(age_seq, dlco_m, type = "l", lwd = 2, col = "#235787",
     xlab = "Age (years)", ylab = "Predicted DLCOc (mmol/min/kPa)",
     main = "Kainu et al. (2017) -- Predicted DLCOc vs age",
     ylim = range(c(dlco_m, dlco_f)))
lines(age_seq, dlco_f, lwd = 2, col = "#e63946")
legend("topright", legend = c("Male 177 cm, 80 kg", "Female 164 cm, 65 kg"),
       col = c("#235787", "#e63946"), lwd = 2, bty = "n")
```

### DLCO z-score verification

```{r dlco_zscore_check}
pred_dlco <- dlco_pred(50, 177, 80, 1, "DLCO")
z_dlco <- dlco_zscore(50, 177, 80, 1, DLCO = pred_dlco)
cat("Predicted DLCOc:", round(pred_dlco, 4), "mmol/min/kPa\n")
cat("Z-score when observed = predicted:", round(z_dlco, 6), "\n")
```

### DLCO lower limit of normal

```{r dlco_lln}
dlco_lln_table <- data.frame(
  age = ages,
  pred_DLCO = round(dlco_pred(ages, 177, 80, 1, "DLCO"), 2),
  LLN_DLCO = round(dlco_lln(ages, 177, 80, 1, "DLCO"), 2),
  pred_VA = round(dlco_pred(ages, 177, 80, 1, "VA"), 2),
  LLN_VA = round(dlco_lln(ages, 177, 80, 1, "VA"), 2)
)
knitr::kable(dlco_lln_table, caption = "Predicted vs LLN -- males, 177 cm, 80 kg")
```
