---
title: "Validating rkainu against Kainu et al. (2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validating rkainu against Kainu et al. (2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rkainu)
```

## Reference

Kainu A, Timonen KL, Toikka J, et al. Reference values of spirometry for
Finnish adults. *Clin Physiol Funct Imaging*. 2016;36(5):346-358.

The `rkainu` package implements the reference equations from this paper,
covering 10 spirometry parameters for Finnish adults aged 18--90.

## Validation against published values

We compare `kainu_pred()` output against expected values computed directly
from the published equations and supplement data.

### Males, 177 cm

```{r males}
ages <- c(20, 30, 40, 50, 60, 70, 80)
params <- c("FVC", "FEV1", "FEV1FVC", "PEF")

male_pred <- kainu_pred(ages, 177, 1, params)
male_pred <- cbind(age = ages, round(male_pred, 3))
knitr::kable(male_pred, caption = "Predicted values — males, 177 cm")
```

### Females, 164 cm

```{r females}
female_pred <- kainu_pred(ages, 164, 2, params)
female_pred <- cbind(age = ages, round(female_pred, 3))
knitr::kable(female_pred, caption = "Predicted values — females, 164 cm")
```

## FEV1 decline with age

```{r fev1_plot, fig.width=6, fig.height=4}
age_seq <- seq(18, 84, by = 1)
fev1_m <- kainu_pred(age_seq, 177, 1, "FEV1")
fev1_f <- kainu_pred(age_seq, 164, 2, "FEV1")

plot(age_seq, fev1_m, type = "l", lwd = 2, col = "#235787",
     xlab = "Age (years)", ylab = "Predicted FEV1 (L)",
     main = "Kainu et al. (2015) — Predicted FEV1 vs age",
     ylim = c(1.5, 5.5))
lines(age_seq, fev1_f, lwd = 2, col = "#e63946")
legend("topright", legend = c("Male 177 cm", "Female 164 cm"),
       col = c("#235787", "#e63946"), lwd = 2, bty = "n")
```

## Z-score verification

When the observed value equals the predicted value, the z-score should be
exactly zero:

```{r zscore_check}
pred_fev1 <- kainu_pred(50, 177, 1, "FEV1")
z <- kainu_zscore(50, 177, 1, FEV1 = pred_fev1)
cat("Predicted FEV1:", round(pred_fev1, 4), "L\n")
cat("Z-score when observed = predicted:", z, "\n")
```

## Lower limit of normal

```{r lln}
lln_table <- data.frame(
  age = ages,
  pred_FEV1 = round(kainu_pred(ages, 177, 1, "FEV1"), 3),
  LLN_FEV1 = round(kainu_lln(ages, 177, 1, "FEV1"), 3),
  pred_FVC = round(kainu_pred(ages, 177, 1, "FVC"), 3),
  LLN_FVC = round(kainu_lln(ages, 177, 1, "FVC"), 3)
)
knitr::kable(lln_table, caption = "Predicted vs LLN — males, 177 cm")
```

## Data.frame workflow

```{r df_example}
patients <- data.frame(
  id = 1:3,
  age = c(35, 55, 72),
  height = c(182, 170, 168),
  sex = c(1, 2, 1),
  FEV1 = c(4.2, 2.5, 2.1),
  FVC = c(5.3, 3.2, 3.0)
)

result <- kainu_df(patients, params = c("FEV1", "FVC"))
knitr::kable(round(result[, -1], 2), caption = "kainu_df() output")
```
